Ainur:~A Framework for Repeatable End-to-End Wireless Edge Computing Testbed Research

### 2022-12-19 publications/2022Ainur/software_layers/cloud.tex
glossaries

Integration with \ac{AWS} is handled by the \mintinline{python}{CloudInstances} class.
The implementation of this class uses \verb|boto3| for interaction with the \ac{AWS}, and thus needs access to appropriate credentials\footnote{%
    \url{https://boto3.amazonaws.com/v1/documentation/api/latest/guide/credentials.html}%
}.

Each instance of this class is associated with a specific \ac{AWS} region, and manages the end-to-end lifecycle of an arbitrary number of \ac{AWS} \ac{EC2} instances on said region.
Specifically, \mintinline{python}{CloudInstances} handles
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item configuring the necessary prerequisites for deploying the instances (security groups and private access keys)
    \item deploying the instances and waiting for them to successfully boot
    \item shutting down and terminating the instances
    \item cleaning up access keys and security groups
\end{enumerate*}.
Instances of this class can be used as context managers, automating this way the termination of instances and cleanup of resources when execution exits the \mintinline{python}{with} block.

By default, \mintinline{python}{CloudInstances} spawns instances of type \verb|t3.micro|, with at least one security group allowing inbound \ac{SSH} traffic and unrestricted traffic between instances.
The type of instances spawned can be modified at runtime, and additional security groups can be added at will.

The spawned \ac{EC2} instances are created from a preconfigured \ac{AMI}, the ID of which must be passed to the \mintinline{python}{CloudInstances} instance at runtime.
This \ac{AMI} can be either be one of the default \acp{AMI} provided by \acp{AWS} --- e.g.\ a plain Ubuntu or CentOS \ac{AMI} --- or a custom \ac{AMI} which has been preconfigured by the user.
Note that although the \mintinline{python}{CloudInstances} instance places no restrictions on the \ac{AMI} used for deploying (with the exception that it needs to be a Linux-based \ac{AMI}), components such as the VPN layer (see \cref{sec:vpn}) depending on these cloud instances might have requirements of their own.

>>>>

Integration with \gls{AWS} is handled by the \mintinline{python}{CloudInstances} class.
The implementation of this class uses \verb|boto3| for interaction with the \gls{AWS}, and thus needs access to appropriate credentials\footnote{%
    \url{https://boto3.amazonaws.com/v1/documentation/api/latest/guide/credentials.html}%
}.

Each instance of this class is associated with a specific \gls{AWS} region, and manages the end-to-end lifecycle of an arbitrary number of \gls{AWS} \gls{EC2} instances on said region.
Specifically, \mintinline{python}{CloudInstances} handles
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item configuring the necessary prerequisites for deploying the instances (security groups and private access keys)
    \item deploying the instances and waiting for them to successfully boot
    \item shutting down and terminating the instances
    \item cleaning up access keys and security groups
\end{enumerate*}.
Instances of this class can be used as context managers, automating this way the termination of instances and cleanup of resources when execution exits the \mintinline{python}{with} block.

By default, \mintinline{python}{CloudInstances} spawns instances of type \verb|t3.micro|, with at least one security group allowing inbound \gls{SSH} traffic and unrestricted traffic between instances.
The type of instances spawned can be modified at runtime, and additional security groups can be added at will.

The spawned \gls{EC2} instances are created from a preconfigured \gls{AMI}, the ID of which must be passed to the \mintinline{python}{CloudInstances} instance at runtime.
This \gls{AMI} can be either be one of the default \acp{AMI} provided by \acp{AWS} --- e.g.\ a plain Ubuntu or CentOS \gls{AMI} --- or a custom \gls{AMI} which has been preconfigured by the user.
Note that although the \mintinline{python}{CloudInstances} instance places no restrictions on the \gls{AMI} used for deploying (with the exception that it needs to be a Linux-based \gls{AMI}), components such as the VPN layer (see \cref{sec:vpn}) depending on these cloud instances might have requirements of their own.

### 2022-12-19 publications/2022Ainur/software_layers/flow.tex
glossaries

\Cref{fig:flow} illustrates the execution flow of Ainur.
The software follows a mostly linear approach given its direct interaction with distinct layers of hardware and networking.
Immediately after start, Ainur reads and parses the deployment configuration and parses it into a set of various dataclasses and Python dictionaries.
Next, if deploying anything to the cloud, Ainur spawns the required number of \ac{AWS} \ac{EC2} instance.
Execution then goes on to configure the physical links, connecting hosts with each other through clever management of \acp{VLAN} and \acp{SDR}.
Once physical and/or \ac{VLAN} connections are configure, \ac{IP} addresses are assigned to devices and static routes between them are configured.
If cloud instances are used, \ac{VPN} connections are established to them.
At this stage, the testbed network is fully configured, and bidirectional traffic between any two hosts is possible over the workload data network.

>>>>

\Cref{fig:flow} illustrates the execution flow of Ainur.
The software follows a mostly linear approach given its direct interaction with distinct layers of hardware and networking.
Immediately after start, Ainur reads and parses the deployment configuration and parses it into a set of various dataclasses and Python dictionaries.
Next, if deploying anything to the cloud, Ainur spawns the required number of \gls{AWS} \gls{EC2} instance.
Execution then goes on to configure the physical links, connecting hosts with each other through clever management of \acp{VLAN} and \acp{SDR}.
Once physical and/or \gls{VLAN} connections are configure, \gls{IP} addresses are assigned to devices and static routes between them are configured.
If cloud instances are used, \gls{VPN} connections are established to them.
At this stage, the testbed network is fully configured, and bidirectional traffic between any two hosts is possible over the workload data network.

### 2022-12-19 publications/2022Ainur/software_layers/ip_layer.tex
glossaries

\subsubsection{\ac{IP} Layer Connectivity}\label{sec:layer3}

Layer 3 connectivity, that is, \ac{IP} addressing and routing, is handled by two main components in Ainur: \mintinline{python}{LANLayer} and \mintinline{python}{VPNCloudMesh}.

\mintinline{python}{LANLayer} handles local network connectivity.
It can be used as a context manager for automatic network teardown, as well as an iterator over the hosts it manages.
This class exposes two main \ac{API} methods, \mintinline{python}{self.add_hosts()} and \mintinline{python}{self.tear_down()}:
\begin{description}[]
    \item[\mintinline{python}{LANLayer.add_hosts(self, layer2)}]
    This method takes a \mintinline{python}{PhysicalLayer} object instance and configures \ac{IP} layer connectivity on the hosts within.
    This includes assigning \ac{IP} addresses to interfaces, configuring default and static routes, as well as authenticating and connecting to the desired access point in the case of WiFi.
    The underlying implementation uses Ansible to write Netplan YAML configuration files on the target hosts and subsequently apply the network configuration.

    \item[\mintinline{python}{LANLayer.tear_down(self)}]
    Performs a clean teardown of \ac{IP} layer connectivity of the hosts managed by the instance, removing the created Netplan configuration files and reverting host network configuration to its initial state.
    When \mintinline{python}{LANLayer} is used as a context manager, this method is automatically called the moment execution leaves the \mintinline{python}{with} block.
\end{description}

\mintinline{python}{VPNCloudMesh}\label{sec:vpn} handles \ac{VPN} connectivity to running cloud instances.
It makes some strong assumptions about the presence of two preconfigured \ac{VPN} gateways on the testbed (some of these have been mentioned above, but we will repeat them here for clarity):

\begin{itemize}
    \item Two VPNCloud gateways, one for control connections and one for workload data streams, are configured and accessible on the control network.
    \item One of these gateways is additionally connected to the workload data network through a separate interface.
    \item These gateways are both listening on separate \ac{UDP} ports on a single public IP address.
    In other words, both gateways access the public internet through a single \ac{NAT} gateway, on which appropriate \ac{UDP} port forwarding rules are configured to allow \ac{VPN} traffic back into the control network.
    \item The VPNCloud connections configured on the the gateways share the same pre-shared key.
    \item The VPNCloud links belong to mutually \emph{disjoint} \acs{IP}v4 address ranges.
    % \item The gateways are assigned the first host address in the available address space of their respective \ac{VPN} links.
    % For example, if the gateway \ac{VPN} link address space is \texttt{10.0.0.0/24}, the gateway has address \texttt{10.0.0.1} on the \ac{VPN} link.
    \item Devices on the control network have been preconfigured with appropriate \ac{IP} routes to reach the control gateway and its associated \ac{VPN} link.
    \item Finally, cloud instances intended to be used with this class have VPNCloud installed on them.
\end{itemize}

@@

The \ac{API} of this class exposes the below relevant methods:

@@

    \begin{description}[font=\normalfont\ttfamily\underline]
        \item[gateway\_ip] An \texttt{IPv4Address} object corresponding to the shared public \ac{IP}v4 address of the \ac{VPN} gateways (i.e.\ the public \ac{IP} address of the \ac{NAT} gateway.)

        \item[vpn\_psk]
        The pre-shared key used to authenticate and encrypt traffic on the \ac{VPN} links.

@@

        \item[gw\_mgmt\_ip~\normalfont{and} gw\_wkld\_ip]
        Full \ac{IP} addresses of the \ac{VPN} links for the control and workload data networks, respectively, including subnet masks.
        These values are passed as \mintinline[breakafter=4]{python}{IPv4Interface} objects, and are used to set up correct routing tables to the cloud instances.

@@

        \item[mgmt\_port~\normalfont{and} wkld\_port]
        \ac{UDP} ports on the \ac{NAT} gateway forwarded to the control and workload gateways, respectively.
    \end{description}

@@

    The full signature of this method is \mintinline[breakafter=.,breakafter=_]{python}{VPNCloudMesh.connect_cloud(self, cloud_layer, host_configs)}.
    The \texttt{cloud\_layer} argument corresponds to a \mintinline{python}{CloudInstances} object instance, and the \texttt{host\_configs} to a collection of host configuration dataclasses.
    This method matches the host configurations to cloud instances in the \mintinline{python}{CloudInstances} object and connects them to to the local control and workload data networks through the two aforementioned separate \ac{VPN} links.
    The underlying implementation uses a combination of the \texttt{boto3} library and Ansible to configure appropriate security groups on the instances and subsequently initialize and connect the \ac{VPN} links.

    \item[\mintinline{python}{VPNCloudMesh.tear_down(self)}]
    This method tears down the \ac{VPN} links, disconnecting the remote instances from the gateways and removing any created routes to the cloud from local devices.
    When an instance of this class is used as a context manager, this method is automatically called at the end of the \mintinline{python}{with} block, ensuring automatic cleanup of resources.

>>>>

\subsubsection{\gls{IP} Layer Connectivity}\label{sec:layer3}

Layer 3 connectivity, that is, \gls{IP} addressing and routing, is handled by two main components in Ainur: \mintinline{python}{LANLayer} and \mintinline{python}{VPNCloudMesh}.

\mintinline{python}{LANLayer} handles local network connectivity.
It can be used as a context manager for automatic network teardown, as well as an iterator over the hosts it manages.
This class exposes two main \gls{API} methods, \mintinline{python}{self.add_hosts()} and \mintinline{python}{self.tear_down()}:
\begin{description}[]
    \item[\mintinline{python}{LANLayer.add_hosts(self, layer2)}]
    This method takes a \mintinline{python}{PhysicalLayer} object instance and configures \gls{IP} layer connectivity on the hosts within.
    This includes assigning \gls{IP} addresses to interfaces, configuring default and static routes, as well as authenticating and connecting to the desired access point in the case of WiFi.
    The underlying implementation uses Ansible to write Netplan YAML configuration files on the target hosts and subsequently apply the network configuration.

    \item[\mintinline{python}{LANLayer.tear_down(self)}]
    Performs a clean teardown of \gls{IP} layer connectivity of the hosts managed by the instance, removing the created Netplan configuration files and reverting host network configuration to its initial state.
    When \mintinline{python}{LANLayer} is used as a context manager, this method is automatically called the moment execution leaves the \mintinline{python}{with} block.
\end{description}

\mintinline{python}{VPNCloudMesh}\label{sec:vpn} handles \gls{VPN} connectivity to running cloud instances.
It makes some strong assumptions about the presence of two preconfigured \gls{VPN} gateways on the testbed (some of these have been mentioned above, but we will repeat them here for clarity):

\begin{itemize}
    \item Two VPNCloud gateways, one for control connections and one for workload data streams, are configured and accessible on the control network.
    \item One of these gateways is additionally connected to the workload data network through a separate interface.
    \item These gateways are both listening on separate \gls{UDP} ports on a single public IP address.
    In other words, both gateways access the public internet through a single \gls{NAT} gateway, on which appropriate \gls{UDP} port forwarding rules are configured to allow \gls{VPN} traffic back into the control network.
    \item The VPNCloud connections configured on the the gateways share the same pre-shared key.
    \item The VPNCloud links belong to mutually \emph{disjoint} \acs{IP}v4 address ranges.
    % \item The gateways are assigned the first host address in the available address space of their respective \gls{VPN} links.
    % For example, if the gateway \gls{VPN} link address space is \texttt{10.0.0.0/24}, the gateway has address \texttt{10.0.0.1} on the \gls{VPN} link.
    \item Devices on the control network have been preconfigured with appropriate \gls{IP} routes to reach the control gateway and its associated \gls{VPN} link.
    \item Finally, cloud instances intended to be used with this class have VPNCloud installed on them.
\end{itemize}

@@

The \gls{API} of this class exposes the below relevant methods:

@@

    \begin{description}[font=\normalfont\ttfamily\underline]
        \item[gateway\_ip] An \texttt{IPv4Address} object corresponding to the shared public \gls{IP}v4 address of the \gls{VPN} gateways (i.e.\ the public \gls{IP} address of the \gls{NAT} gateway.)

        \item[vpn\_psk]
        The pre-shared key used to authenticate and encrypt traffic on the \gls{VPN} links.

@@

        \item[gw\_mgmt\_ip~\normalfont{and} gw\_wkld\_ip]
        Full \gls{IP} addresses of the \gls{VPN} links for the control and workload data networks, respectively, including subnet masks.
        These values are passed as \mintinline[breakafter=4]{python}{IPv4Interface} objects, and are used to set up correct routing tables to the cloud instances.

@@

        \item[mgmt\_port~\normalfont{and} wkld\_port]
        \gls{UDP} ports on the \gls{NAT} gateway forwarded to the control and workload gateways, respectively.
    \end{description}

@@

    The full signature of this method is \mintinline[breakafter=.,breakafter=_]{python}{VPNCloudMesh.connect_cloud(self, cloud_layer, host_configs)}.
    The \texttt{cloud\_layer} argument corresponds to a \mintinline{python}{CloudInstances} object instance, and the \texttt{host\_configs} to a collection of host configuration dataclasses.
    This method matches the host configurations to cloud instances in the \mintinline{python}{CloudInstances} object and connects them to to the local control and workload data networks through the two aforementioned separate \gls{VPN} links.
    The underlying implementation uses a combination of the \texttt{boto3} library and Ansible to configure appropriate security groups on the instances and subsequently initialize and connect the \gls{VPN} links.

    \item[\mintinline{python}{VPNCloudMesh.tear_down(self)}]
    This method tears down the \gls{VPN} links, disconnecting the remote instances from the gateways and removing any created routes to the cloud from local devices.
    When an instance of this class is used as a context manager, this method is automatically called at the end of the \mintinline{python}{with} block, ensuring automatic cleanup of resources.

### 2022-12-19 publications/2022Ainur/software_layers/physical.tex
glossaries

This layer creates and deploy the underlying physical connections of the data network and is implemented in the \mintinline{python}{PhysicalLayer} class. 
Types of connections handled by this layer could be ethernet and WiFi links, or even an entire software-defined 4G \ac{LTE} network.

@@

Currently, WiFi and \ac{LTE} implementations are available. 
User-defined implementation for other wireless connections are possible and encouraged.

@@

    \mintinline{python}{StationSoftwareDefinedRadio} class is used for the station WiFi nodes.
    It searches for an WiFi \ac{SSID} among the networks that are being advertised and attempts for establishing the connection.

@@

    \item[\ac{SDR}-based \ac{LTE} Network]
    \mintinline{python}{SDRLTE} class can configure and deploy a complete software-defined \ac{LTE} network.
    The constructor must be provided with an \mintinline{python}{EPC} class, a set of \mintinline{python}{ENodeB} classes, and a set of \mintinline{python}{LTEUE} classes from the \emph{autoran} Python package\footnote{See https://github.com/samiemostafavi/autoran}.
    This package exploits the \emph{docker-py} framework to run containerized OpenAirInterface \ac{LTE} functions on remote hosts.
    Network configurations are needed in the form of Python dictionaries for the construction of \emph{autoran} classes.

    \ac{LTE} \ac{EPC} consists of HSS, MME, SPGW-U, and SPGW-C services which can be run on the same host.
    For each \ac{eNodeB} or \ac{UE}, a software-defined radio and a host for signal processing must be specified additionally.
    The resulting topology of an \ac{LTE} network in Ainur is depicted in Figure \ref{fig:lte}.
    Unlike WiFi wireless connection, due to the fact that \ac{LTE} creates an internal ip network and that introduces routing complexities, the server host and the client host cannot be on the same subnet in the ip layer.

    \begin{figure}[t]
        \centering
        \includegraphics[width=0.8\linewidth]{figures/lte.png}
        \caption{An example of \ac{LTE} data network topology in data network and its corresponding \acp{VLAN}}
        \label{fig:lte}
    \end{figure}

>>>>

This layer creates and deploy the underlying physical connections of the data network and is implemented in the \mintinline{python}{PhysicalLayer} class. 
Types of connections handled by this layer could be ethernet and WiFi links, or even an entire software-defined 4G \gls{LTE} network.

@@

Currently, WiFi and \gls{LTE} implementations are available.
User-defined implementation for other wireless connections are possible and encouraged.

@@

    \mintinline{python}{StationSoftwareDefinedRadio} class is used for the station WiFi nodes.
    It searches for an WiFi \gls{SSID} among the networks that are being advertised and attempts for establishing the connection.

@@

    \item[\gls{SDR}-based \gls{LTE} Network]
    \mintinline{python}{SDRLTE} class can configure and deploy a complete software-defined \gls{LTE} network.
    The constructor must be provided with an \mintinline{python}{EPC} class, a set of \mintinline{python}{ENodeB} classes, and a set of \mintinline{python}{LTEUE} classes from the \emph{autoran} Python package\footnote{See https://github.com/samiemostafavi/autoran}.
    This package exploits the \emph{docker-py} framework to run containerized OpenAirInterface \gls{LTE} functions on remote hosts.
    Network configurations are needed in the form of Python dictionaries for the construction of \emph{autoran} classes.

    \gls{LTE} \gls{EPC} consists of HSS, MME, SPGW-U, and SPGW-C services which can be run on the same host.
    For each \gls{eNodeB} or \gls{UE}, a software-defined radio and a host for signal processing must be specified additionally.
    The resulting topology of an \gls{LTE} network in Ainur is depicted in Figure \ref{fig:lte}.
    Unlike WiFi wireless connection, due to the fact that \gls{LTE} creates an internal ip network and that introduces routing complexities, the server host and the client host cannot be on the same subnet in the ip layer.

    \begin{figure}[t]
        \centering
        \includegraphics[width=0.8\linewidth]{figures/lte.png}
        \caption{An example of \gls{LTE} data network topology in data network and its corresponding \acp{VLAN}}
        \label{fig:lte}
    \end{figure}

### 2022-12-19 publications/2022Ainur/software_layers/storage_logging.tex
glossaries

Dynamic, distributed storage shared across testbed hosts is managed by the \mintinline{python}{ExperimentStorage} class.
This class exposes a very simple \ac{API}:
\begin{itemize}
    \item At instantiation time, the instance deploys a samba\footnote{\url{https://www.samba.org/}} server container on a specified host, using a host folder mounted inside the container for persistent storage.
    Then, on each host configured on the workload data network, it configures a Docker Volume\footnote{\url{https://docs.docker.com/storage/volumes/}} pointing to the samba server.
    These Docker Volumes are only locally available, but since they all share the same identifier, they can be included in deployment configurations and mounted on applications orchestrated with Docker Swarm (see \cref{sec:swarm} below).

@@

Logging of the experiment is carried out using a containerized approach.
This is achieved through the combination of \texttt{docker-py} and \textit{FluentBit} -- an open-source and multi-platform log processor and forwarding tool.
At deployment time, Ainur configures a FluentBit client container in each workload host which captures logs produced by the workload.
These FluentBit clients are configured to listen on a \ac{TCP} port to which workload logs are forwarded by Dockers built-in FluentD logging driver\footnote{\url{https://docs.docker.com/config/containers/logging/fluentd/}}.
This means that logs are automatically captured by Docker from the \emph{standard output} of the containerized workload process.
The FluentBit client containers append metadata such as timestamp, workload name, workload host ID, to the collected logs and forward them to a central FluentBit or FluentD server.

>>>>

Dynamic, distributed storage shared across testbed hosts is managed by the \mintinline{python}{ExperimentStorage} class.
This class exposes a very simple \gls{API}:
\begin{itemize}
    \item At instantiation time, the instance deploys a samba\footnote{\url{https://www.samba.org/}} server container on a specified host, using a host folder mounted inside the container for persistent storage.
    Then, on each host configured on the workload data network, it configures a Docker Volume\footnote{\url{https://docs.docker.com/storage/volumes/}} pointing to the samba server.
    These Docker Volumes are only locally available, but since they all share the same identifier, they can be included in deployment configurations and mounted on applications orchestrated with Docker Swarm (see \cref{sec:swarm} below).

@@

Logging of the experiment is carried out using a containerized approach.
This is achieved through the combination of \texttt{docker-py} and \textit{FluentBit} -- an open-source and multi-platform log processor and forwarding tool.
At deployment time, Ainur configures a FluentBit client container in each workload host which captures logs produced by the workload.
These FluentBit clients are configured to listen on a \gls{TCP} port to which workload logs are forwarded by Dockers built-in FluentD logging driver\footnote{\url{https://docs.docker.com/config/containers/logging/fluentd/}}.
This means that logs are automatically captured by Docker from the \emph{standard output} of the containerized workload process.
The FluentBit client containers append metadata such as timestamp, workload name, workload host ID, to the collected logs and forward them to a central FluentBit or FluentD server.

### 2022-12-19 publications/2022Ainur/software_layers/swarm.tex
glossaries

A simple example of the usage of this class and its basic \ac{API} can be observed in \cref{lst:swarm:example}, and in the following we will detail the major steps in operating this class:
\begin{description}[]
    \item[Instantiating a \mintinline{python}{DockerSwarm} object.]
    Obtaining an instance of a Swarm object is straightforward, as the constructor takes no parameters.
    \item[Adding Swarm nodes.]
    Swarm manager and worker nodes are added through the \mintinline{python}{deploy_managers()} and \mintinline{python}{deploy_workers()} methods respectively.
    These methods take a single mandatory parameter \mintinline{python}{hosts}, corresponding a mapping from \mintinline{python}{AinurHost} instances to dictionaries of node variables in the form \mintinline[breaklines, breakafter={_}]{python}{{'label_name': 'label_value'}}.
    Additional keyword arguments are parsed as default values for labels which are assigned to all nodes.

>>>>

A simple example of the usage of this class and its basic \gls{API} can be observed in \cref{lst:swarm:example}, and in the following we will detail the major steps in operating this class:
\begin{description}[]
    \item[Instantiating a \mintinline{python}{DockerSwarm} object.]
    Obtaining an instance of a Swarm object is straightforward, as the constructor takes no parameters.
    \item[Adding Swarm nodes.]
    Swarm manager and worker nodes are added through the \mintinline{python}{deploy_managers()} and \mintinline{python}{deploy_workers()} methods respectively.
    These methods take a single mandatory parameter \mintinline{python}{hosts}, corresponding a mapping from \mintinline{python}{AinurHost} instances to dictionaries of node variables in the form \mintinline[breaklines, breakafter={_}]{python}{{'label_name': 'label_value'}}.
    Additional keyword arguments are parsed as default values for labels which are assigned to all nodes.
